---
title: "Hands-on-Exercise 1: A Layered Grammer of Graphics"
author: "Kartikey Thakar"
Date: "April 10, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview


In this hands-on exercise, I explore the foundations of data visualisation 
in R using **ggplot2**, which is built on the principles of the 
*Layered Grammar of Graphics*. The goal is to understand how each 
grammatical layer contributes to building meaningful statistical graphics, 
and to gain hands-on experience creating a variety of chart types.

---

## Getting Started

### Loading the Libraries

To get started, I will use the `pacman` package to load `tidyverse`, 
which includes `ggplot2` and other essential data wrangling tools. 
Using `p_load()` is convenient as it installs the package if it isn't 
already installed, then loads it.

```{r}
pacman::p_load(tidyverse)
```

### Importing the Data

Next, I import the dataset using `read_csv()` from the `readr` package 
(part of tidyverse). The dataset contains year-end examination scores 
for a cohort of Primary 3 students, with the following variables:

- **Categorical**: ID, CLASS, GENDER, RACE
- **Continuous**: ENGLISH, MATHS, SCIENCE

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

---

## Introducing ggplot2

### R Graphics vs ggplot2

Before diving into ggplot2, let me first compare it with Base R's 
built-in graphics to understand why ggplot2 is preferred.

Using Base R:

```{r}
hist(exam_data$MATHS)
```

Now the same histogram using ggplot2:

```{r}
ggplot(data = exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 10,
                 boundary = 100,
                 color = "black",
                 fill = "grey") +
  ggtitle("Distribution of Maths scores")
```

While the Base R version is simpler to write, ggplot2 gives me much 
more control and flexibility over the visual output. More importantly, 
it encourages thinking about visualisation as a mapping between 
data variables and visual properties — a skill that transfers across 
all kinds of charts.

---

## Grammar of Graphics

ggplot2 is built on Leland Wilkinson's Grammar of Graphics, which 
breaks a graphic down into seven distinct layers:

| Layer | Description |
|---|---|
| **Data** | The dataset being plotted |
| **Aesthetics** | Maps variables to visual properties like colour, size, position |
| **Geometrics** | The visual marks — bars, points, lines, etc. |
| **Facets** | Splits data into small multiples for comparison |
| **Statistics** | Summarises data (e.g. means, confidence intervals) |
| **Coordinates** | Defines the plotting plane and axis system |
| **Themes** | Controls non-data styling like fonts and backgrounds |

Let me now work through each of these layers one by one.

---

## Essential Element 1: Data

I start by calling `ggplot()` with just the dataset. This gives me 
a blank canvas — ggplot2 knows what data to use, but hasn't been 
told what to draw yet.

```{r}
ggplot(data = exam_data)
```

As expected, nothing appears yet except an empty plot area.

---

## Essential Element 2: Aesthetic Mappings

Next, I add an aesthetic mapping using `aes()` to tell ggplot2 
which variable to place on the x-axis. This adds the axis and 
its label, but still no geometric shapes yet.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS))
```

---

## Essential Element 3: Geometric Objects

Now things get interesting. Geometric objects (geoms) are the actual 
visual marks placed on the plot. I will go through several geom types 
using the exam dataset.

### geom_bar

Let me start with a simple bar chart to visualise the count of 
students by race.

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

### geom_dotplot

A dot plot is another way to visualise distribution, where each dot 
represents one observation. I also turn off the y-axis here since 
it is not meaningful in this context.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_dotplot(binwidth = 2.5,
               dotsize = 0.5) +
  scale_y_continuous(NULL,
                     breaks = NULL)
```

### geom_histogram

A histogram is one of the most common ways to visualise the 
distribution of a continuous variable. Here I plot the distribution 
of Maths scores with 20 bins.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```

I can also map a categorical variable like GENDER to the fill 
aesthetic to see the distribution broken down by gender.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           fill = GENDER)) +
  geom_histogram(bins = 20,
                 color = "grey30")
```

### geom_density

A kernel density plot is a smoothed alternative to the histogram, 
useful for continuous data from an underlying smooth distribution.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_density()
```

By mapping GENDER to the colour aesthetic, I can compare the 
distributions of Maths scores between male and female students.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           colour = GENDER)) +
  geom_density()
```

### geom_boxplot

Box plots are great for visualising the five-number summary 
(minimum, Q1, median, Q3, maximum) and identifying outliers. 
Here I compare Maths scores across genders.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot()
```

I can also add notches to the box plot to help assess whether 
the medians between groups are statistically different. 
Non-overlapping notches suggest the medians differ.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot(notch = TRUE)
```

### geom_violin

Violin plots combine a box plot with a density plot, making it 
easier to compare distributions across groups visually.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_violin()
```

### geom_point

Scatter plots are useful for exploring relationships between 
two continuous variables. Here I look at whether Maths and 
English scores are correlated.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point()
```

### Combining Geoms

One of the strengths of ggplot2 is that geoms can be layered 
on top of each other. Here I overlay individual data points 
on a box plot using `geom_point()` with jitter to avoid overplotting.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot() +
  geom_point(position = "jitter",
             size = 0.5)
```

---

## Essential Element 4: Statistics

Statistical transformations let me summarise data directly within 
the plot. Here I add the mean score as a red point on top of 
the box plots using `stat_summary()`.

```{r}
ggplot(data = exam_data,
       aes(y = MATHS, x = GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",
               fun = "mean",
               colour = "red",
               size = 4)
```

I can also use `geom_smooth()` to add a best-fit curve to a 
scatter plot. The default method is loess (locally weighted smoothing).

```{r}
ggplot(data = exam_data,
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(linewidth = 0.5)
```

Alternatively, I can switch to a linear regression line by 
specifying `method = lm`.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5)
```

---

## Essential Element 5: Facets

Facetting allows me to split the data into subsets and display 
each as a separate panel — great for comparing distributions 
across groups.

### facet_wrap

Here I use `facet_wrap()` to plot a histogram of Maths scores 
for each class, arranged in a 2D grid automatically.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ CLASS)
```

### facet_grid

`facet_grid()` lays panels out in a strict row-column matrix. 
Below I use it to display histograms across classes in a 
single row.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20) +
  facet_grid(~ CLASS)
```

---

## Essential Element 6: Coordinates

Coordinate systems control how the axes are oriented and scaled.

### coord_flip

By default, bar charts are vertical. I can flip the axes using 
`coord_flip()` to make it horizontal, which is often more 
readable when category labels are long.

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar() +
  coord_flip()
```

### coord_cartesian

When comparing Maths and English scores on a scatter plot, 
both axes should have the same range for a fair visual 
comparison. I use `coord_cartesian()` to fix both axes to 0–100.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))
```

---

## Essential Element 7: Themes

Themes let me change the overall look and feel of the plot 
without touching the data. Let me compare three built-in themes.

`theme_gray()` — the default ggplot2 theme:

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

`theme_classic()` — a cleaner look with no gridlines:

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

`theme_minimal()` — minimal styling, my personal preference 
for clean academic outputs:

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

---

## References

- Hadley Wickham (2023) *ggplot2: Elegant Graphics for Data Analysis*. 
  Online 3rd edition.
- Winston Chang (2013) *R Graphics Cookbook*, 2nd edition.
- Healy, Kieran (2019) *Data Visualization: A practical introduction*.
